Evaluating Cultural Transmission in Bronze Age burial rites of Central, Northern and North-western Europe using radiocarbon data

### 2019-05-26 article/02_Materials_and_Methods.Rmd
reflecting the changes in Figure 6 in the text

Figure \ref{correlationtimeseries} shows the development of the correlation coefficient values obtained via Mantel statistics (Table \ref{tab:dacta} & \ref{tab:dactb}) during the succession of the seven relevant 200 years time slots.

>>>>

Figure \ref{correlationtimeseries} B shows the development of the correlation coefficient values obtained via Mantel statistics (Table \ref{tab:dacta} & \ref{tab:dactb}) during the succession of the seven relevant 200 years time slots. Because of the highly suggestive nature of this figure, it is complemented by Figure \ref{correlationtimeseries} A, which shows the (very low!) median number of graves available for each region and year to calculate these correlations.

### 2019-05-26 article/03_Results.Rmd
reflecting the changes in Figure 6 in the text

Table \ref{tab:dactb} shows that the distance networks of *burial type* and *burial construction* are not equivalent. There is also no significant negative correlation. The degree of agreement between two regions in the question of inhumation and cremation is therefore not suitable for predicting the similarity in the dominance of flat or mound graves. Only for the time slot at the transition to the Middle Bronze Age from 1800 to 1600 calBC the 5% significance level of the Mantel test is almost undercut, thus suggesting a statistically significant relationship. This moment of strong correlation is though not induced by similarity along the spatial dimension: The effect is still visible in the result of the Partial Mantel test, where the spatial distance is artificially counterbalanced (see also Figure \ref{correlationtimeseries}).

For the relationship between the SED networks and the spatial distance network several observations result from Table \ref{tab:dacta}:

@@

Looking at the SED network of *burial construction*, a slightly different conclusions can be drawn: The distance values here are larger overall, and the SED differences within the individual spatial distance classes are clearly more accentuated. As well as in *burial type* the Early Bronze Age is by far the phase with the most substantial spatial correlation, though it is only significant for the time slot between 2000 and 1400 calBC. There is no time period with negative mean correlation for *burial construction* cultural distance and spatial distance, but also here the correlation coefficient approaches zero in the Late Bronze Age.

In general, cultural distance defined here does not seem to correlate strongly with spatial distance, especially in the Middle and Late Bronze Age.

>>>>

Table \ref{tab:dactb} and Figure \ref{correlationtimeseries} show that the distance networks of *burial type* and *burial construction* are not equivalent. There is also no significant negative correlation. The degree of agreement between two regions in the question of inhumation and cremation is therefore not suitable for predicting the similarity in the dominance of flat or mound graves. Only for the time slot at the transition to the Middle Bronze Age from 1800 to 1600 calBC the 5% significance level of the Mantel test is almost undercut, thus suggesting a statistically significant relationship. This moment of strong correlation is though not induced by similarity along the spatial dimension alone: The effect is still partially visible in the result of the Partial Mantel test, where the spatial distance is artificially counterbalanced.

For the relationship between the SED networks and the spatial distance network several observations result from Table \ref{tab:dacta} and Figure \ref{correlationtimeseries}:

@@

Looking at the SED network of *burial construction*, a slightly different conclusions can be drawn: The distance values here are larger, and the SED differences within the individual spatial distance classes are clearly more accentuated. As well as in *burial type* the Early Bronze Age is by far the phase with the most substantial spatial correlation. But it is only significant for the time slot between 2000 and 1400 calBC. There is no time period with negative mean correlation for *burial construction* cultural distance and spatial distance, but also here the correlation coefficient approaches zero in the Late Bronze Age. These latter results in particular, however, are based on unsatisfactory few observations.

In general, cultural distance as defined here does not seem to correlate strongly with spatial distance, especially in the Middle and Late Bronze Age. However, this hypothesis is not conclusive at the moment as it is derived from a very small sample size.

### 2019-05-26 article/04_Discussion.Rmd
reflecting the changes in Figure 6 in the text

The second important question posed for this paper was, *how cultural distance between regions develops, and how it is related to spatial distance?* While the idea behind the first question was rather technical -- reproducing already available information in a quantitative way -- the second one had a larger scope. The quantitative representation of an archaeological narrative paths the way for a diverse set of methods to be applied and the analysis undertaken here is merely the beginning.

@@

Against this background, the observation that there is no permanent, significant positive correlation either between the variables or with spatial distance is all the more astonishing (Table \ref{tab:dacta} and \ref{tab:dactb}, Figure \ref{correlationtimeseries}). Only for certain periods of time this correlation is visible in the Radon-B data. The *burial construction* distance generally shows a higher degree of correlation with spatial distance than the *burial type* distance, which even seems to behave in an opposing manner: negative correlation (also not statistically significant!). If one assumes the reconstructed timelines of cultural similarity intensity to be correct, than one has to explain a rise (between 2200 and 2000 calBC) and a decline (after 2000 calBC) of cultural-spatial-distance correlation. There are several reasons why any interpretation of this observation is difficult so far:

@@

Keeping this in mind, the surprising result is still valid and has to be explained. One interpretation might be, that the Early Bronze Age was a time of rising cultural exchange and religious homogenization, while in later periods different domains secluded themselves more strongly. The negative correlation of *burial type* in the Late Bronze Age could indicate long distance contacts that explicitly excluded certain intermediate regions of Central Europe. The rise of the Tumulus Culture in the Middle Bronze Age might have cushioned this effect for the variable *burial construction* and introduced more variability for a -- at that time -- less important aspect *burial type*. The latter may have been of subordinate relevance in the Tumulus culture context. With the rise of the Urnfield culture this phenomenon could have been reversed as *burial type* became an expression of something that could have been one of the first formalized religions in Europe: The spread of the Urnfield Culture was an extraordinary development and possibly accompanied by a structured, religious dogma that could have built up around the symbolic content of cult wagons, sun barks and waterfowl [@bilicSwanChariotSolar2016; @kimmig_seevolkerbewegung_1964; @molloy_european_2018]. Of course these explanations could, if correct, only be considered on a big scale. For individual regions and sites specific ideological, social or even economic considerations might have been much more critical for the decision which burial tradition to adopt [@barrett_monumentality_1990; @fokkens_genesis_1997-1; @muller-scheesel_variabilitat_2009].

>>>>

The second important question posed for this paper was, *how cultural distance between regions develops, and how it is related to spatial distance?* While the idea behind the first question was rather technical -- reproducing already available information in a quantitative way -- the second one had a larger scope. The quantitative representation of an archaeological narrative paths the way for a diverse set of methods to be applied and the analysis undertaken here is merely the beginning. Unfortunately, the small sample size makes it very difficult to obtain meaningful, significant results in this context.

@@

Against this background, the observation that there does not seem to be permanent, significant positive correlation neither in between the variables nor with spatial distance is all the more astonishing (Table \ref{tab:dacta} and \ref{tab:dactb}, Figure \ref{correlationtimeseries} B). Only for certain periods of time this correlation is visible in the Radon-B data. The *burial construction* distance generally shows a higher degree of correlation with spatial distance than the *burial type* distance, which even seems to behave in an opposing manner: negative correlation (not statistically significant!). If one assumes the reconstructed timelines of cultural similarity intensity to be correct, than one has to explain a rise (between 2200 and 2000 calBC) and a decline (after 2000 calBC) of cultural-spatial-distance correlation. There are several reasons why any interpretation of this observation is difficult so far:

@@

Keeping this in mind, the surprising result still can be discussed. One interpretation might be, that the Early Bronze Age was a time of rising cultural exchange and religious homogenization, while in later periods different domains secluded themselves more strongly. The negative correlation of *burial type* in the Late Bronze Age could indicate long distance contacts that explicitly excluded certain intermediate regions of Central Europe. The rise of the Tumulus Culture in the Middle Bronze Age might have cushioned this effect for the variable *burial construction* and introduced more variability for a -- at that time -- less important aspect *burial type*. The latter may have been of subordinate relevance in the Tumulus culture context. With the rise of the Urnfield culture this phenomenon could have been reversed as *burial type* became an expression of something that could have been one of the first formalized religions in Europe: The spread of the Urnfield Culture was an extraordinary development and possibly accompanied by a structured, religious dogma built up around the symbolic content of cult wagons, sun barks and waterfowl [@bilicSwanChariotSolar2016; @kimmig_seevolkerbewegung_1964; @molloy_european_2018]. Of course these explanations could, if correct, only be considered on a big scale. For individual regions and sites specific ideological, social or even economic considerations might have been much more critical for the decision which burial tradition to adopt [@barrett_monumentality_1990; @fokkens_genesis_1997-1; @muller-scheesel_variabilitat_2009].

### 2019-05-26 article/05_Conclusion.Rmd
reflecting the changes in Figure 6 in the text

Even small amounts of radiocarbon dates can be transformed to meaningful regional time series, if they can be linked to the relevant meta information. Although only `r r("gpr size", sf_desc)` graves -- a small fraction of maybe hundreds of thousands burials that can be estimated for the European Bronze Age [@nikulkaArchaologischeDemographieMethoden2016] -- have been available for this analysis, the resulting reconstruction of general burial tradition trends in wast areas of Central, Northern and North-western Europe comes close to the one achieved by literature research. The time series can be employed for cultural distance calculation. They yield the surprising observation that cultural distance and spatial distance do not significantly correlate. Although this interpretation must be treated with caution, it seems as if the spread of burial traditions are not simple *Diffusion of Innovation* events as introduced by @rogersDiffusionInnovations1962, but the result of more complex interactions between competing ideas.

>>>>

Even small amounts of radiocarbon dates can be transformed to meaningful regional time series, if they can be linked to the relevant meta information. Although only `r r("gpr size", sf_desc)` graves -- a small fraction of maybe hundreds of thousands burials that can be estimated for the European Bronze Age [@nikulkaArchaologischeDemographieMethoden2016] -- have been available for this analysis, the resulting reconstruction of general burial tradition trends in wast areas of Central, Northern and North-western Europe comes close to the one achieved by literature research. The time series can be employed for cultural distance calculation. They yield the surprising observation that cultural distance and spatial distance do not significantly correlate. Although this result must be treated with caution because of the very small sample size, it seems as if the spread of burial traditions are not simple *Diffusion of Innovation* events as introduced by @rogersDiffusionInnovations1962, but the result of more complex interactions between competing ideas.
